Hydrodynamics of a Semipermeable Inextensible Membrane Under Flow and Confinement

### 2021-02-11 vesicle_code/docs/permeableVes/semipermeable_vesicle.tex
updated time stepping section

In our formulation, a semipermeable vesicle is suspended in a viscous Newtonian fluid 
(in domain $\Omega$).  The vesicle membrane $\gamma$ is parameterized as $\xx(s,t)$
where $s$ is arclength and $t$ is time. In the suspending fluid the flow
velocity $\uu$ and pressure $p$ are governed by the incompressible 
Stokes equations
\begin{align}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu = 0, \quad
  \nabla \cdot \uu = 0, \qquad \xx \in \Omega \setminus \gamma,
\end{align}
where $\mu$ is the fluid viscosity. When $\Omega$ is an unconfined
geometry, the condition $\uu(\xx) \rightarrow \uu_\infty(\xx)$ as $|\xx|
\rightarrow \infty$ is enforced. 
%
Along the vesicle membrane $\gamma$,
mass continuity, force balance, and local inextensibility are enforced
by requiring
\begin{align}
  \label{eqn:jump}
  \jump{\uu} = 0, \quad
  \jump{T\cdot\nn} = \ff_\mathrm{mem}, \quad
  \xx_s \cdot \uu_s = 0,
\end{align}
where 
%$\nn$ is the outward normal of $\gamma$, 
$\jump{\cdot}$ is the jump across the interface, and $T$ is the
hydrodynamic stress tensor.
\todo[inline]{We defined $T$ to be the pressure in equation (1)}
%The membrane force $\ff_\mathrm{mem}$ is the sum of a bending force
%with bending modulus $k_b$ and a force due to the tension $\Lambda$.
%Following~\citet{vee-gue-zor-bir2009}, we introduce the variable
%$\sigma = \Lambda - \frac{3}{2}\kappa^2$, where $\kappa$ is the vesicle
%curvature. Then, the bending and tension forces corresponding to the
%Helfrich energy are $\ff_\mathrm{ben} = -k_b \xx_{ssss}$ and
%$\ff_\mathrm{ten} = (\sigma \xx_s)_s$.

@@

Since the fluid satisfies the Stokes equations, the velocity and
pressure can be represented as layer potentials. Given the interfacial
boundary conditions in Equation~\eqref{eqn:jump}, the fluid velocity is
\begin{align}
  \uu(\xx) = \uu_\infty(\xx) + \SS[\ff_\mathrm{mem}](\xx), \quad
    \xx \in \Omega,
\end{align}
where $\SS$ is the single-layer potential
\begin{align}
  \SS[\ff](\xx) = \frac{1}{4\pi} \int_{\gamma} \left(
    -\mathbf{I} \log\rho + \frac{\rr \otimes \rr}{\rho^2} \right)
    \ff(\yy) ds_{\yy},
%  \SS[\ff](\xx) = \frac{1}{4\pi\mu} \int_{\gamma} \left(
%    -\mathbf{I} \log\rho + \frac{\rr \otimes \rr}{\rho^2} \right)
%    \ff(\yy) ds_{\yy},
\end{align}
with $\rr = \xx - \yy$ and $\rho = |\rr|$. 
Together with the inextensibility constraint and the dimensionless boundary condition
\begin{align}
  \label{eqn:vesicleFlux}
  \uu - \dot{\xx} = - \beta (\ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
%
%Semipermeability is introduced by modifying the vesicle velocity from
%the standard no-slip condition $\dot{\xx} = \uu(\xx)$. Instead,
%the permeating water flux through the lipid bilayer membrane is proportional to the
%membrane traction~\cite{yao-mor2017}. That is, 
%\begin{align}
%  \label{eqn:vesicleFlux}
%  \uu - \dot{\xx} = - \beta (\ff_\mathrm{mem} \cdot \nn) \nn, \qquad
%  \xx \in \gamma,
%\end{align}
%where $\beta = k_w/(R_0/\mu)$ is the dimensionless membrane permeability for water.
%
%Imposing the flux
%condition~\eqref{eqn:vesicleFlux} and the inextensibility condition, 
the vesicle membrane velocity satisfies the boundary integral equation
\begin{align}
  \label{eqn:vesVelocity}
  \dot{\xx} &= \uu_\infty(\xx) + \beta (\ff_\mathrm{mem}\cdot\nn)\nn
  + \SS [\ff_\mathrm{mem}](\xx),  \quad
  \xx_s \cdot \dot{\xx}_s = 0.
\end{align}
In confined geometries, the far field term $\uu_\infty(\xx)$ is replaced
by a double-layer potential with an unknown density defined on
$\partial\Omega$. The density function is used to satisfy the boundary
condition on $\partial\Omega$. Specifically, in the two examples of a semipermeable vesicle under strong confinement, we enforce a
no-slip boundary condition along the top and bottom of the channel and a
Hagen–Poiseuille flow at the inlet and outlet---this boundary condition
has been used in similar vesicle setups~\cite{qua-bir2014,
rah-vee-bir2010, lu-rah-zor2017}.

@@

\subsection{Discretization in Time}
To eliminate stringent time step restrictions, we modify a time stepping
method that discretizes high-order derivatives
semi-implicitly~\cite{vee-gue-zor-bir2009, qua-bir2014}. We introduce
the notation
\begin{alignat}{3}
  &B[\xx]\ff = -\frac{d^4}{ds^4} \ff,  \qquad
  &&T[\xx]\sigma = (\sigma \xx_s)_s, \\
  &D[\xx]\ff = \xx_s \cdot \ff_s, 
  &&P[\xx]\ff = (\ff \cdot \nn) \nn,
\end{alignat}
for the differential and projection operators. Then,
Equation~\eqref{eqn:vesVelocity} is written as
\begin{align}
  \dot{\xx} &= \uu_\infty(\xx) + \beta P(B\xx + T\sigma)
  + \SS (B\xx + T\sigma), \quad
  D \dot{\xx} = 0.
\end{align}
We note that the integral and differential operators, which all depend
on the vesicle shape, are non-linear with respect to the vesicle
shape. We construct a time stepping scheme by letting $B^N$ be the
bending operator due to the vesicle configuration at time $t^N$, and use
similar notation for the other operators. Then, a first-order
semi-implicit time stepping method of Equation~\eqref{eqn:vesVelocity} is
\begin{alignat}{3}  
  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right),  \\
%  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
%  &&+ \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) \nonumber \\
%  & &&+ \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right),  \\
  D^N\xx^{N+1} &= 1,
\end{alignat}
and this linear system is solved with matrix-free GMRES.

>>>>

In our formulation, a semipermeable vesicle is suspended in a viscous
Newtonian fluid (in domain $\Omega$).  The vesicle membrane $\gamma$ is
parameterized as $\xx(s,t)$ where $s$ is arclength and $t$ is time. In
the suspending fluid the flow velocity $\uu$ and pressure $p$ are
governed by the incompressible Stokes equations
\begin{align}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu = 0, \quad
  \nabla \cdot \uu = 0, \qquad \xx \in \Omega \setminus \gamma,
\end{align}
where $\mu$ is the fluid viscosity. When $\Omega$ is an unconfined
geometry, the condition $\uu(\xx) \rightarrow \uu_\infty(\xx)$ as $|\xx|
\rightarrow \infty$ is enforced. 
%
Along the vesicle membrane $\gamma$,
mass continuity, force balance, and local inextensibility are enforced
by requiring
\begin{align}
  \label{eqn:jump}
  \jump{\uu} = 0, \quad
  \jump{\mathbf{T}\cdot\nn} = \ff_\mathrm{mem}, \quad
  \xx_s \cdot \uu_s = 0,
\end{align}
where 
%$\nn$ is the outward normal of $\gamma$, 
$\jump{\cdot}$ is the jump across the interface, and $\mathbf{T} = -p
\mathbf{I} + \mu \left(\nabla \uu + \nabla \uu^T \right)$ is the
hydrodynamic stress tensor.
%The membrane force $\ff_\mathrm{mem}$ is the sum of a bending force
%with bending modulus $k_b$ and a force due to the tension $\Lambda$.
%Following~\citet{vee-gue-zor-bir2009}, we introduce the variable
%$\sigma = \Lambda - \frac{3}{2}\kappa^2$, where $\kappa$ is the vesicle
%curvature. Then, the bending and tension forces corresponding to the
%Helfrich energy are $\ff_\mathrm{ben} = -k_b \xx_{ssss}$ and
%$\ff_\mathrm{ten} = (\sigma \xx_s)_s$.

@@

Since the fluid satisfies the Stokes equations, the velocity and
pressure can be represented as layer potentials. Given the interfacial
boundary conditions in Equation~\eqref{eqn:jump}, the fluid velocity is
\begin{align}
  \uu(\xx) = \uu_\infty(\xx) + \SS[\ff_\mathrm{mem}](\xx), \quad
    \xx \in \Omega,
\end{align}
where $\SS$ is the single-layer potential
\begin{align}
  \label{eqn:SLP}
  \SS[\ff](\xx) = \frac{1}{4\pi} \int_{\gamma} \left(
    -\mathbf{I} \log\rho + \frac{\rr \otimes \rr}{\rho^2} \right)
    \ff(\yy) ds_{\yy},
%  \SS[\ff](\xx) = \frac{1}{4\pi\mu} \int_{\gamma} \left(
%    -\mathbf{I} \log\rho + \frac{\rr \otimes \rr}{\rho^2} \right)
%    \ff(\yy) ds_{\yy},
\end{align}
with $\rr = \xx - \yy$ and $\rho = |\rr|$. 
Together with the inextensibility constraint and the dimensionless boundary condition
\begin{align}
  \label{eqn:vesicleFlux}
  \uu - \dot{\xx} = - \beta (\ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
%
%Semipermeability is introduced by modifying the vesicle velocity from
%the standard no-slip condition $\dot{\xx} = \uu(\xx)$. Instead,
%the permeating water flux through the lipid bilayer membrane is proportional to the
%membrane traction~\cite{yao-mor2017}. That is, 
%\begin{align}
%  \label{eqn:vesicleFlux}
%  \uu - \dot{\xx} = - \beta (\ff_\mathrm{mem} \cdot \nn) \nn, \qquad
%  \xx \in \gamma,
%\end{align}
%where $\beta = k_w/(R_0/\mu)$ is the dimensionless membrane permeability for water.
%
%Imposing the flux
%condition~\eqref{eqn:vesicleFlux} and the inextensibility condition, 
the vesicle membrane velocity satisfies the boundary integral equation
\begin{align}
  \label{eqn:vesVelocity}
  \dot{\xx} &= \uu_\infty(\xx) + \beta (\ff_\mathrm{mem}\cdot\nn)\nn
  + \SS [\ff_\mathrm{mem}](\xx),  \quad
  \xx_s \cdot \dot{\xx}_s = 0.
\end{align}
In confined geometries, the far field term $\uu_\infty(\xx)$ is replaced
by a double-layer potential with an unknown density defined on
$\partial\Omega$. The density function is used to satisfy the boundary
condition on $\partial\Omega$. Specifically, in the two examples of a semipermeable vesicle under strong confinement, we enforce a
no-slip boundary condition along the top and bottom of the channel and a
Hagen–Poiseuille flow at the inlet and outlet---this boundary condition
has been used in similar vesicle setups~\cite{qua-bir2014,
rah-vee-bir2010, lu-rah-zor2017}.

@@

\subsection{Discretization in Time}
To eliminate stringent time step restrictions, we modify a time stepping
method that discretizes high-order derivatives
semi-implicitly~\cite{vee-gue-zor-bir2009, qua-bir2014}. We introduce
the notation
%\begin{alignat}{3}
%  &\BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
%  &&\TT\sigma = (\sigma \xx_s)_s, \\
%  &\DD\ff = \xx_s \cdot \ff_s, 
%  &&P\ff = (\ff \cdot \nn) \nn,
%\end{alignat}
\begin{align}
  \BB\ff = -\frac{d^4}{ds^4} \ff,  \qquad
  \TT\sigma = (\sigma \xx_s)_s, \quad
  \DD\ff = \xx_s \cdot \ff_s, \quad
  P\ff = (\ff \cdot \nn) \nn,
  \label{eqn:operators}
\end{align}
for the differential and projection operators. Then,
Equation~\eqref{eqn:vesVelocity} can be written as
\begin{align}
  \dot{\xx} &= \uu_\infty(\xx) + \beta P(\BB\xx + \TT\sigma)
  + \SS (\BB\xx + \TT\sigma), \quad \DD \dot{\xx} = 0.
\end{align}
We note that the operators defined in Equation~\eqref{eqn:operators}, as
well as the single-layer potential defined in Equation~\eqref{eqn:SLP}
all depend on the vesicle shape. A time stepping method that treats
these dependencies implicitly would result in a non-linear set of
equations that would be computationally expensive to solve. Instead, we
use a time stepping method that defines the operators at time $t^N$ and
discretizes the stiff terms implicitly. In particular, we use the
first-order semi-implicit method
%We note that the integral and differential operators, which all depend
%on the vesicle shape, are non-linear with respect to the vesicle shape.
%We construct a time stepping scheme by letting $\BB^N$ be the bending
%operator due to the vesicle configuration at time $t^N$, and use
%similar notation for the other operators. Then, a first-order
%semi-implicit time stepping method of Equation~\eqref{eqn:vesVelocity}
%is
\begin{alignat}{3}  
  \frac{\xx^{N+1} - \xx^N}{\triangle t} &= \uu_\infty(\xx^N) 
  + \beta P^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right) 
  + \SS^N\left(\BB^N\xx^{N+1} + \TT^N\sigma^{N+1}\right),  \\
  \DD^N\xx^{N+1} &= 1,
\end{alignat}
and this linear system is solved with matrix-free GMRES.
